Real-time Attention Assessment for Social Robots

### 2015-09-15 paper.tex
Minor formatting

Successfully sustaining face-to-face interaction between humans and robots
mandates fine perception of each other attentional focus. Along this line, we
present a novel technique to assess the visual focus of attention (VFOA) based
on accurate, real-time head pose estimation with monocular camera. The software
is fully open-source and can readily be adopted on a wide range of social
robots. We propose an in-depth evaluation of the technique, using manual video
annotations as ground truth, and show that our system reaches high level of
accuracy in a real-world face-to-face interaction scenario. We finally suggest how
to derive a metric of engagement from the monitoring of the VFOA, allowing the
robot to account and react to the user's changing engagement in real-time.

@@

Engagement describes a cognitive, affective (specifically intrinsic motivation),
and behavioral state of interaction with a computer application that ``makes the
user want to be there" \cite{OBrien:2010}. Engaging interactions were thought to
involve intrinsic interest, interactivity, perceived control and choice,
functionality, motivation and attention. In fact,
\cite{yarbus1967eye,barber1976perception} suggest that human's attention is
highly correlated with what they look at, and the existing relationship between
gaze and attention during social interaction has been exposed
in~\cite{argyle1969social}. In this way, in situ Visual Focus of Attention
(VFOA) estimation is advantageous to human-robot interaction and a major
contribution towards a better understanding of the interaction quality.

@@

In this research, we address the problem of tracking the children's engagement
based on their VFOA and current performed task during a long term child-robot
interaction in a learning context, first. And second, we provide qualitative and
quantitative evaluation of the outcome provided using a first person RGB sensor
approach. Moreover, this monitoring activity has been supporting our research in
child-robot interaction for learning and instruction purposes.

@@

\paragraph{Real-Time Assessment of Visual Focus of Attention}
The participants' VFOA acquisition is the only dependent variable. This includes
their gaze times and patterns of looking or not looking to the targets. This
data extracted from an RGB camera was also coded post-hoc to quantify some of
these measurements.

We predict that there will be no generalizable pattern of participant's VFOA
behavior for different scenarios such as multiple subjects, necessitating in
situ VFOA estimation capabilities to enable robots to adapt in the specific
scenarios. However, we believe that off-line pattern identification can lead to
a better understanding of the children progression as well as engagement level
assessment in face-to-face long term interactions.\fixme{I did not understand
this paragraph}

@@

We propose a adaptive model based solution to approach head pose estimation that
finds correspondences between facial landmarks points and their respective
locations in a face model. We use the open-source library
dlib\footnote{http://dlib.net} \cite{dlib09}, to detect and extract the skeleton
of the subject's face and then, use the obtained landmarkers to calculate the
head pose orientation estimation (see figure \ref{system}). In this way, we can
estimate what the subject was looking at and obtain an \textit{rviz} 3D scene
representation as it is executed as a ROS (Robot Operating System) node that can
be seen in figure \ref{rviz}. Compared to intrusive hardware like head mounted
displays, eye-trackers or wearables for instance, this approach has a clear
advantage of simplicity and integration with robotic systems, but also becomes
accurate enough for general purposes in face-to-face interactions. In addition,
it allows us to model complex environments, handling different number and
positions of participants in the scene.

\begin{figure}[h!]
    \centering
    %\includegraphics[width=0.9\columnwidth]{system}
    \caption{\small System overview showing the most relevant components involved in the estimation.}
    \label{system}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=0.9\columnwidth]{rviz_camera}
    \caption{\small Rviz screenshot showing Nao, camera (Nao's feet), face detected with its field of view (gray cone) and blue marker intersecting as current VFOA.}
    \label{rviz}
\end{figure}

@@

Assuming that the \textit{tf} frames representing the robotic agent and the VFOA
candidates are static, need to be monitored. This task consists on evaluating
the intersection between the \textit{tf} frames and the field of view: If one
focus of attention is within the region defined by the field of view, we assume
the subject is looking at it. In order to solve such operation mathematically it
is necessary to compute the transformation matrix between the coordinate frames
to be able to locate the monitored frames \textit{A} in the subject's coordinate
system \textit{B} (see equation \ref{eq:transform}).

@@

Explain that: other modules have to provide the sitation assessment (what
is where?) in terms of published frames, and a last module compute the
intersection of the VFOA and these frames\fixme{TBD}

@@

The results has proved that the VFOA prediction accuracy in the tracking is not
only conditioned by the subject's amount of movement, but also by the speed of
the same movements: Restless children externalize greater levels of movement
decreasing the VFOA prediction accuracy.

For instance, subjects 1,4 and 5 despite the fact that the tracking accuracy is
in the average, the bad precision in the captured VFOA suggests that such
children are highly likely to modify the gaze direction without head move
intervention reducing dramatically the overlap between captured and real VFOA.

@@

\begin{figure*}
    \centering
    \includegraphics[width=1.8\columnwidth]{realExpected}
    \caption{\small In blue, real VFOA of subject 4 accoding to two observer's
    assessment. In cyan, the matching between the real and the obtained VFOA. In
    yellow, the repeated pattern in turn-taking and in orange, the experimenter's
    suggestion to choose a new word.}

@@

    % Maybe we should add the expected VFOA
\end{figure*}

\begin{figure*}
    \centering
    \includegraphics[width=1.81\columnwidth]{realCaptured}
    \caption{\small Real VFOA of subject 4 accoding to two observer's
    assessment. In green, the obtained VFOA using our aproach and in cyan, the
    mathing between them. Finally, the correspondent numbers with figure \ref{fig:bitmap}.}

@@

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.7\columnwidth]{bitmap}
    \caption{\small Cosine similarity matrix between real and cap-tured VFOA
    using a 10 sec. window size computation. The correspondent numbers are shown
    in figure \ref{fig:realCaptured}.}
    \label{fig:bitmap}
\end{figure}

@@

Additional information of the head pose can be obtained for the evaluation of
the VFOA distribution. Figure \ref{fig:heatmap} shows the most salient focuses
of attention where the peaks are located. These peaks are mainly projected
between the triangle composed by the robot, the writing tablet and the selection
tablet using an unsupervised K-means clustering method reveals, considering the
observers as negligible VFOA.

% Workaround to do subfigure since the packege looks not to work
\begin{figure}
  \centering
  \begin{tabular}{@{}c@{}}
    \includegraphics[width=.7\linewidth,height=100pt]{heatmap} \\
    \small (a)  2D histogram.
  \end{tabular}

@@

In this work we have presented an on-line vision-based engagement assessment
solution that can be extended to work in different platforms for simulating
complex interaction scenarios monitoring several VFOA with a maker-less
technology. It provides some benefits like the decrement of the system
complexity as well as the increment of the naturalness of the interaction due to
the use of an RGB sensor.

@@

Regarding to the approach used, some limitations need to be considered. The most
relevant one, the fact that the positions of the VFOA need to be known in
advance and defined as static \textit{tf} frames in the system. Moreover, it is
important to notice that eye gaze information is neglected which induces a
certain amount of error, as well as the noise induced by the 3D pose estimation
computed from face features.

@@

Gaze information acquired during the interactions is not only useful to detect
and predict engagement, but also allows the possibility to develop actively the
concept gaze-aware robot. In other words, make the system able to estimate where
the subject looks at, but also imitate the same subject's VFOA when the activity
state allows it.

>>>>

Successfully sustaining face-to-face interaction between humans and robots
mandates fine perception of each other attentional focus. Along this line, we
present a novel technique to assess the visual focus of attention (VFoA) based
on accurate, real-time head pose estimation with monocular camera. The software
is fully open-source and can readily be adopted on a wide range of social
robots. We propose an in-depth evaluation of the technique, using manual video
annotations as ground truth, and show that our system reaches high level of
accuracy in a real-world face-to-face interaction scenario. We finally suggest how
to derive a metric of engagement from the monitoring of the VFoA, allowing the
robot to account and react to the user's changing engagement in real-time.

@@

Engagement describes a cognitive, affective (specifically intrinsic motivation),
and behavioral state of interaction with a computer application that ``makes the
user want to be there" \cite{OBrien:2010}. Engaging interactions were thought to
involve intrinsic interest, interactivity, perceived control and choice,
functionality, motivation and attention. In fact,
\cite{yarbus1967eye,barber1976perception} suggest that human's attention is
highly correlated with what they look at, and the existing relationship between
gaze and attention during social interaction has been exposed
in~\cite{argyle1969social}. In this way, in situ Visual Focus of Attention
(VFoA) estimation is advantageous to human-robot interaction and a major
contribution towards a better understanding of the interaction quality.

@@

In this research, we address the problem of tracking the children's engagement
based on their VFoA and current performed task during a long term child-robot
interaction in a learning context, first. And second, we provide qualitative and
quantitative evaluation of the outcome provided using a first person RGB sensor
approach. Moreover, this monitoring activity has been supporting our research in
child-robot interaction for learning and instruction purposes.

@@

\paragraph{Real-Time Assessment of Visual Focus of Attention}

The participants' VFoA acquisition is the only dependent variable. This includes
their gaze times and patterns of looking or not looking to the targets. This
data extracted from an RGB camera was also coded post-hoc to quantify some of
these measurements.

We predict that there will be no generalizable pattern of participant's VFoA
behavior for different scenarios such as multiple subjects, necessitating in
situ VFoA estimation capabilities to enable robots to adapt in the specific
scenarios. However, we believe that off-line pattern identification can lead to
a better understanding of the children progression as well as engagement level
assessment in face-to-face long term interactions.\fixme{I did not understand
this paragraph}

@@

We propose a adaptive model based solution to approach head pose estimation that
finds correspondences between facial landmarks points and their respective
locations in a face model. We use the open-source library
dlib\footnote{http://dlib.net} \cite{dlib09}, to detect and extract the skeleton
of the subject's face and then, use the obtained features to calculate the
head pose orientation estimation (see figure \ref{system}). In this way, we can
estimate what the subject was looking at and obtain an \textit{rviz} 3D scene
representation as it is executed as a ROS (Robot Operating System) node that can
be seen in figure \ref{rviz}. Compared to intrusive hardware like head mounted
displays, eye-trackers or wearables for instance, this approach has a clear
advantage of simplicity and integration with robotic systems, but also becomes
accurate enough for general purposes in face-to-face interactions. In addition,
it allows us to model complex environments, handling different number and
positions of participants in the scene.

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.9\columnwidth]{system}
    \caption{\small System overview showing the most relevant components involved in the estimation.}
    \label{system}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=0.9\columnwidth]{rviz_camera}
    \caption{\small Rviz screenshot showing Nao, camera (Nao's feet), face detected with its field of view (gray cone) and blue marker intersecting as current VFoA.}
    \label{rviz}
\end{figure}

@@

Assuming that the \textit{tf} frames representing the robotic agent and the VFoA
candidates are static, need to be monitored. This task consists on evaluating
the intersection between the \textit{tf} frames and the field of view: If one
focus of attention is within the region defined by the field of view, we assume
the subject is looking at it. In order to solve such operation mathematically it
is necessary to compute the transformation matrix between the coordinate frames
to be able to locate the monitored frames \textit{A} in the subject's coordinate
system \textit{B} (see equation \ref{eq:transform}).

@@

Explain that: other modules have to provide the sitation assessment (what
is where?) in terms of published frames, and a last module compute the
intersection of the VFoA and these frames\fixme{TBD}

@@

The results has proved that the VFoA prediction accuracy in the tracking is not
only conditioned by the subject's amount of movement, but also by the speed of
the same movements: Restless children externalize greater levels of movement
decreasing the VFoA prediction accuracy.

For instance, subjects 1,4 and 5 despite the fact that the tracking accuracy is
in the average, the bad precision in the captured VFoA suggests that such
children are highly likely to modify the gaze direction without head move
intervention reducing dramatically the overlap between captured and real VFoA.

@@

\begin{figure*}
    \centering
    \includegraphics[width=1.8\columnwidth]{realExpected}
    \caption{\small In blue, real VFoA of subject 4 accoding to two observer's
    assessment. In cyan, the matching between the real and the obtained VFoA. In
    yellow, the repeated pattern in turn-taking and in orange, the experimenter's
    suggestion to choose a new word.}

@@

    % Maybe we should add the expected VFoA
\end{figure*}

\begin{figure*}
    \centering
    \includegraphics[width=1.81\columnwidth]{realCaptured}
    \caption{\small Real VFoA of subject 4 accoding to two observer's
    assessment. In green, the obtained VFoA using our aproach and in cyan, the
    mathing between them. Finally, the correspondent numbers with figure \ref{fig:bitmap}.}

@@

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.7\columnwidth]{bitmap}
    \caption{\small Cosine similarity matrix between real and cap-tured VFoA
    using a 10 sec. window size computation. The correspondent numbers are shown
    in figure \ref{fig:realCaptured}.}
    \label{fig:bitmap}
\end{figure}

@@

Additional information of the head pose can be obtained for the evaluation of
the VFoA distribution. Figure \ref{fig:heatmap} shows the most salient focuses
of attention where the peaks are located. These peaks are mainly projected
between the triangle composed by the robot, the writing tablet and the selection
tablet using an unsupervised K-means clustering method reveals, considering the
observers as negligible VFoA.

% Workaround to do subfigure since the packege looks not to work
\begin{figure}[th!]
  \centering
  \begin{tabular}{@{}c@{}}
    \includegraphics[width=.7\linewidth,height=100pt]{heatmap} \\
    \small (a)  2D histogram.
  \end{tabular}

@@

In this work we have presented an on-line vision-based engagement assessment
solution that can be extended to work in different platforms for simulating
complex interaction scenarios monitoring several VFoA with a maker-less
technology. It provides some benefits like the decrement of the system
complexity as well as the increment of the naturalness of the interaction due to
the use of an RGB sensor.

@@

Regarding to the approach used, some limitations need to be considered. The most
relevant one, the fact that the positions of the VFoA need to be known in
advance and defined as static \textit{tf} frames in the system. Moreover, it is
important to notice that eye gaze information is neglected which induces a
certain amount of error, as well as the noise induced by the 3D pose estimation
computed from face features.

@@

Gaze information acquired during the interactions is not only useful to detect
and predict engagement, but also allows the possibility to develop actively the
concept gaze-aware robot. In other words, make the system able to estimate where
the subject looks at, but also imitate the same subject's VFoA when the activity
state allows it.
